---
title: "R lab assignment 7: Markov models"
author: "Keegan Ballantyne"
output: html_document
---

Before starting this assignment read chapter 13 and work through the first part of R tutorial 10.


## Simulations of a two state model of infection

In this section you will simulate a two-state Markov model, by starting with a value of the state variable, using a random number and an if statement to generate the next value, and then the next one, and so on.  Since Markov models are inherently random, each simulation will result in a different sequence of states.

Consider a simple model of an infectious epidemic, in which the population consists of only two types of people: susceptible (S) and infected (I). Suppose that the transition probabilities are (per day): 0.2 (from S to I) and 0.5 (from I to S). The first task asks you to write a script to decide, based on the initial state of the person (S or I), what state they are in the next day. Task 1.2 asks you to repeat that decision (using a for loop) to generate a string of states (S and I) over a given number of days, which you will save as a vector. You will then count the number of days the person spent in each state (S or I) and plot the histogram using the functions `barplot()` and `table()`. Since this is a random simulation, you will get different strings of states every time your run your code.

1. Write a script to a make a random decision for an individual based on the SI Markov model, where you assign the initial state to be either 1 (S) or 2 (I). The script should first use an if statement to check whether the state is 1 or else is 2 and then use the transition probabilities given above to make the random decision. Run this script a few times to make sure it transitions from both state 1 to 2, and from 2 to 1 (remember that it's a random decision, so don't worry if the transition doesn't happen as often as you'd expect) and report the results of your experimentation.

```{r}
set.seed(1)
state <- 1 # set initial state
trans1to2 <- 0.2 # transition probability from 1 to 2
trans2to1 <- 0.5 # transition probability from 2 to 1
decider <- runif(1) # random number between 0 and 1
if (state==1) {
    if (decider < trans1to2) { # randomly decide to transition
        state <- 2
    } else { # or to stay
        state <-1
    }
} else if (state==2) {
    if (decider < trans2to1) { # randomly decide to transition
        state <- 1 
    } else { # or to stay
        state <-2
    }
} else {
    print ("Initial state must be either 1 or 2!")
}
print(state) 
```

Running the code 10 times setting the state as state 1 the code would go from state 1 to state 2 and from state 2 to state 1.
Running the code 10 times setting the state as state 2 the code would go from state 2 to state 1 and from state 1 to state 2.

2. Add a for loop around your previous script to generate a state string over 30 days for an individual given an an initial state. Instead of a state variable preallocate a vector and assign the states for each day inside the loop. Turn this script into a function that takes the following inputs: transition probabilities from 1 to 2 and from 2 to 1, initial state, and number of time steps. The function should return the vector of states. Run this function and plot the resulting vector of states using plot(state.vec) to visualize the simulated disease history of one person and plot the histogram of state frequencies using the command barplot(table(state.vec)). Re-run the script ten times (use a for loop) to see different disease histories and report the fraction of days each of these "people" spend infected.

```{r}
state <- 1 # set initial state
trans1to2 <- 0.2 # transition probability from 1 to 2
trans2to1 <- 0.5 # transition probability from 2 to 1
num <- 30
func <- function(trans1to2, trans2to1, state, num){
  state_vector <- rep(state, 30)
  for(i in 1:num){
     decider <- runif(1) # random number between 0 and 1
    if (state_vector[i]==1) {
    if (decider < trans1to2) { # randomly decide to transition
        state_vector[i+1] <- 2
    } else { # or to stay
        state_vector[i+1] <- 1
    }
    } else if (state_vector[i]==2) {
    if (decider < trans2to1) { # randomly decide to transition
        state_vector[i+1] <- 1 
    } else { # or to stay
        state_vector[i+1] <- 2
    }
} else {
    print ("Initial state must be either 1 or 2!")
}
}
 return(state_vector)
}

for (i in 1:10) {
  state_vector <- func(trans1to2, trans2to1, state, num)
  plot(state_vector, main = "Disease History over 30 Days", xlab = "Time (Days)", ylab = "State of the Individual")
  barplot(table(state_vector), main = "Frequency Histogram", xlab = "States", ylim = c(0,30))
  print(paste("Fraction of Infected Days", table(state_vector)[2]/num))
}
```

Loop 1: "Fraction of Infected Days 0.233333333333333"
Loop 2: "Fraction of Infected Days 0.466666666666667"
Loop 3: "Fraction of Infected Days 0.2"
Loop 4: "Fraction of Infected Days 0.233333333333333"
Loop 5: "Fraction of Infected Days 0.366666666666667"
Loop 6: "Fraction of Infected Days 0.233333333333333"
Loop 7: "Fraction of Infected Days 0.633333333333333"
Loop 8: "Fraction of Infected Days 0.366666666666667"
Loop 9: "Fraction of Infected Days 0.233333333333333"
Loop 10: "Fraction of Infected Days 0.533333333333333"

3. Now change the infection rate (transition from S to I) to 0.05, keep the recovery rate (transition from I to S) of 0.5, and use your function above (no need to redefine it!) to generate a state vector ten times to see different disease histories and report the fraction of days these "people" spend infected. How does it compare to the fractions in the previous task?

```{r}
state <- 1 # set initial state
trans1to2 <- 0.05 # transition probability from 1 to 2
trans2to1 <- 0.5 # transition probability from 2 to 1
num <- 30
func <- function(trans1to2, trans2to1, state, num){
  state_vector <- rep(state, 30)
  for(i in 1:num){
     decider <- runif(1) # random number between 0 and 1
if (state_vector[i]==1) {
    if (decider < trans1to2) { # randomly decide to transition
        state_vector[i+1] <- 2
    } else { # or to stay
        state_vector[i+1] <- 1
    }
} else if (state_vector[i]==2) {
    if (decider < trans2to1) { # randomly decide to transition
        state_vector[i+1] <- 1 
    } else { # or to stay
        state_vector[i+1] <-2
    }
} else {
    print ("Initial state must be either 1 or 2!")
}
}
 return(state_vector)
}

for (i in 1:10) {
  state_vector <- func(trans1to2, trans2to1, state, num)
  plot(state_vector, main = "Disease History over 30 Days", xlab = "Time (Days)", ylab = "State of the Individual")
  barplot(table(state_vector), main = "Frequency Histogram", xlab = "States", ylim = c(0,30))
  print(paste("Fraction of Infected Days", table(state_vector)[2]/num))
}
```

Loop 1: "Fraction of Infected Days 0.0666666666666667"
Loop 2: "Fraction of Infected Days 0.0666666666666667"
Loop 3: "Fraction of Infected Days 0.0666666666666667"
Loop 4: "Fraction of Infected Days 0.166666666666667"
Loop 5: "Fraction of Infected Days 0.0666666666666667"
Loop 6: "Fraction of Infected Days 0.0666666666666667"
Loop 7: "Fraction of Infected Days 0.133333333333333"
Loop 8: "Fraction of Infected Days 0.1"
Loop 9: "Fraction of Infected Days 0.0333333333333333"
Loop 10: "Fraction of Infected Days 0.0333333333333333"

The number of the fraction of infected days is a lot smaller compared to the previous task ranging from 0.03 to 0.13. I think this is due to the fact that the probability of transition of state 1 to state 2 is very small so this means that healthy people tend to stay healthy longer in this data set.

4. Change the recovery rate (transition from I to S) to 0.125, keep the infection rate (transition from S to I) at 0.05 and use your function above (no need to redefine it) to generate a state vector ten times to see different disease histories and report the fraction of days these "people" spend infected. How does it compare to the fractions in the previous two tasks?

```{r}
state <- 1 # set initial state
trans1to2 <- 0.05 # transition probability from 1 to 2
trans2to1 <- 0.125 # transition probability from 2 to 1
num <- 30
func <- function(trans1to2, trans2to1, state, num){
  state_vector <- rep(state, 30)
  for(i in 1:num){
     decider <- runif(1) # random number between 0 and 1
if (state_vector[i]==1) {
    if (decider < trans1to2) { # randomly decide to transition
        state_vector[i+1] <- 2
    } else { # or to stay
        state_vector[i+1] <- 1
    }
} else if (state_vector[i]==2) {
    if (decider < trans2to1) { # randomly decide to transition
        state_vector[i+1] <- 1 
    } else { # or to stay
        state_vector[i+1] <-2
    }
} else {
    print ("Initial state must be either 1 or 2!")
}
}
 return(state_vector)
}

for (i in 1:10) {
  state_vector <- func(trans1to2, trans2to1, state, num)
  plot(state_vector, main = "Disease History over 30 Days", xlab = "Time (Days)", ylab = "State of the Individual")
  barplot(table(state_vector), main = "Frequency Histogram", xlab = "States", ylim = c(0,30))
  print(paste("Fraction of Infected Days", table(state_vector)[2]/num))
}
```

Loop 1: "Fraction of Infected Days 0.166666666666667"
Loop 2: "Fraction of Infected Days 0.6"
Loop 3: "Fraction of Infected Days 0.1"
Loop 4: "Fraction of Infected Days NA"
Loop 5: "Fraction of Infected Days 0.266666666666667"
Loop 6: "Fraction of Infected Days 0.0333333333333333"
Loop 7: "Fraction of Infected Days 0.233333333333333"
Loop 8: "Fraction of Infected Days 0.2"
Loop 9: "Fraction of Infected Days 0.333333333333333"
Loop 10: "Fraction of Infected Days 0.266666666666667"

The number of the fraction of infected days can be really small, like the previous task, or larger, unlike the previous task, or in between ranging from 0.03 to 0.6. I think this is due to the fact that the probability of transition of state 1 to state 2 and the probability of transition of state 2 to state 1 are both small which means that healthy people tend to stay healthy longer (but not as long as infected people staying sick) and infected people tend to stay sick longer in this data set.



